Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-25 agujournaltemplate.tex
added genus level table for gradient forest input data

A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been undertaken on possible changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we utilized here to examine connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data, including X, Y, and Z, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories. Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.

    All in-situ data, inlcuding Niskin bottle, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

@@

\subsection{Data analysis}    
    To provide an overview of the available data and highlight the interannual trends and changes in the time series data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly mean estimates.

    We calculated three metrics of phytoplankton community diversity. First, we calculated genus richness, which reduces taxonomic resolution but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$). 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. 
    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).

@@

Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are from diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high.
Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

@@

>>>>

A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}. The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}. The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

@@

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    Between late 1995 and early 1998, nutrient data, including nitrate, phosphate, and silicate, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate. Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}). A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera. The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.

    All in situ data, including the Niskin bottle, CTD, and phytoplankton taxonomy, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where \mu is the mean across all available data points). Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

    Cell counts of phytoplankton per functional group were assembled per monthly sampling, summed over all individual interpolated counts at the genus or species level, and finally averaged over a year. Nanoflagellate counts were reported as such in the data and not further identified. To characterize phytoplankton community structure, we calculated genus richness, Shannon diversity (H'), and Pielou’s evenness (J') to assess taxonomic variety, community complexity, and the distribution of abundances among genera, respectively. We calculated genus richness as the sum of all identified genera present in a monthly sample. Genus richness provides a lower taxonomic resolution than species richness, but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. The Shannon index was calculated as $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness as $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling. Diversity metrics were calculated for each monthly sample and averaged throughout the year.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped tress and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

@@

\subsection{Interannual changes in the environment and ecosystem}
% Figure 2
Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm). These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix). The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration. Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.  <I think the most obvious pattern in Fig. 2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015. Diversity is high when chl/phyto are low. I think you'll want to describe these patterns as they are interesting.>

@@

% Figure 6
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells. 
Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

@@

%
\begin{table}
\caption{Table of all identified genus that were counted more than 10 times across the time series (n=89). The monthly cell counts per genus were used as input data to the gradient forest analysis.}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}

\begin{tabular}{@{}ll@{}}
\textbf{Diatoms} &                                                            \\
\textit{Amphora}           & C.G. Ehrenberg ex F.T. Kützing, 1844                       \\
\textit{Asterionella}      & A.H. Hassall, 1850                                         \\
\textit{Asterionellopsis}  & F.E. Round in F.E. Round, R.M. Crawford \& D.G. Mann, 1990 \\
\textit{Bacteriastrum}     & G. Shadbolt, 1854                                          \\
\textit{Bellerochea}       & H. Van Heurck, 1885                                        \\
\textit{Cerataulina}       & H. Peragallo ex F. Schütt in Engler \& Prantl, 1896        \\
\textit{Chaetoceros}       & C.G. Ehrenberg, 1844                                       \\
\textit{Climacodium}       & A. Grunow, 1868                                            \\
\textit{Corethron}         & A.F. Castracane, 1886                                      \\
\textit{Coscinodiscus}     & C.G. Ehrenberg, 1839                                       \\
\textit{Cylindrotheca}     & L. Rabenhorst, 1859                                        \\
\textit{Dactyliosolen}     & A.F. Castracane, 1886                                      \\
\textit{Detonula}          & F. Schütt ex G.B. De Toni, 1894                            \\
\textit{Diploneis}         & (C. G. Ehrenberg) P.T. Cleve, 1894                         \\
\textit{Eucampia}          & C.G. Ehrenberg, 1839                                       \\
\textit{Grammatophora}     & C.G. Ehrenberg, 1840                                       \\
\textit{Guinardia}         & H. Peragallo, 1892                                         \\
\textit{Haslea}            & R. Simonsen, 1974                                          \\
\textit{Helicotheca}       & M. Ricard, 1987                                            \\
\textit{Hemiaulus}         & P.A.C. Heiberg, 1863                                       \\
\textit{Lauderia}          & P.T. Cleve, 1873                                           \\
\textit{Leptocylindrus}    & P.T. Cleve in C.G.J. Petersen, 1889                        \\
\textit{Lioloma}           & Hasle, 1997                                                \\
\textit{Navicula}          & J.B.M. Bory de Saint-Vincent, 1822                         \\
\textit{Nitzschia}         & A.H. Hassall, 1845                                         \\
\textit{Odontella}         & C.A. Agardh, 1832                                          \\
\textit{Pinnularia}        & C.G. Ehrenberg, 1843                                       \\
\textit{Pinnunavis}        & H. Okuno, 1975                                             \\
\textit{Planktoniella}     & (C.G.Wallich) Schütt, 1892                                 \\
\textit{Pleurosigma}       & W. Smith, 1852                                             \\
\textit{Proboscia}         & G. Sundström, 1986                                         \\
\textit{Pseudo-nitzschia}  & H. Peragallo in H. Peragallo \& M. Peragallo, 1900         \\
\textit{Rhizosolenia}      & T. Brightwell, 1858                                        \\
\textit{Skeletonema}       & R.K. Greville, 1865                                        \\
\textit{Sundstroemia}      & Medlin, Boonprakob, Lundholm \& Moestrup, 2021             \\
\textit{Tabularia}         & (F.T. Kutzing) D.M. Williams \& F.E. Round, 1986           \\
\textit{Thalassionema}     & A. Grunow ex C. Mereschkowsky, 1902                        \\
\textit{Thalassiosira}     & P.T. Cleve, 1873 emend. Hasle, 1973                        \\
\textbf{Cyanobacteria}     &                                                            \\
\textit{Anabaena}          & Bory de Saint-Vincent ex Bornet \& Flahault, 1886          \\
\textit{Johannesbaptistia} & G.De Toni, 1934                                            \\
\textit{Lyngbya}           & C.Agardh ex Gomont, 1892                                   \\
\textit{Synechococcus}     & Nägeli, 1849                                               \\
\textit{Trichodesmium}     & Ehrenberg ex Gomont, 1892                                  \\
\textbf{Dinoflagellates}    &                                                            \\
\textit{Akashiwo}          & Gert Hansen \& Moestrup, 2000                              \\
\textit{Alexandrium}       & Halim, 1960                                                \\
\textit{Amphidinium}       & Claperède \& Lachmann, 1859                                \\
\textit{Blepharocysta}     & Ehrenberg, 1873                                            \\
\textit{Cochlodinium}      & Schütt, 1896                                               \\
\textit{Cystodinium}       & Klebs, 1912                                                \\

\end{tabular}

\label{tab:sup:GenusFuncGroups}
\end{table}

\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
\begin{tabular}{@{}ll@{}}
\textbf{Dinoflagellates} cont.    &                                                            \\
\textit{Dinophysis}        & Ehrenberg, 1839                                            \\
\textit{Diplopsalis}       & R.S.Bergh, 1881                                            \\
\textit{Glenodinium}       & Ehrenberg, 1836     
\\
\textit{Gonyaulax}         & Diesing, 1866                                              \\
\textit{Gymnodinium}       & F.Stein, 1878                                              \\
\textit{Gyrodinium}        & Kofoid \& Swezy, 1921                                      \\
\textit{Heterocapsa}       & F.Stein, 1883                                              \\
\textit{Lingulodinium}     & D.Wall, 1967                                               \\
\textit{Mesoporos}         & Lillick, 1937                                              \\
\textit{Ornithocercus}     & Stein, 1883                                                \\
\textit{Oxytoxum}          & Stein, 1883                                                \\
\textit{Phalacroma}        & Stein, 1883                                                \\
\textit{Oxyphysis}         & Kofoid, 1926                                               \\
\textit{Podolampas}        & F.Stein, 1883                                              \\
\textit{Pronoctiluca}      & Fabre-Domergue, 1889                                       \\
\textit{Prorocentrum}      & Ehrenberg, 1834                                            \\
\textit{Exuviella}         & Cienkowski, 1881                                           \\
\textit{Protoperidinium}   & Bergh, 1881                                                \\
\textit{Pyrocystis}        & J.Murray ex Haeckel, 1890                                  \\
\textit{Pyrophacus}        & F.Stein, 1883                                              \\
\textit{Scrippsiella}      & Balech ex A.R.Loeblich III, 1965                           \\
\textit{Sphaerodinium}     & Woloszynska, 1916                                          \\
\textit{Triadinium}        & Dodge, 1981                                                \\
\textit{Tripos}            & Bory de Saint-Vincent, 1823                                \\
\textbf{Haptophytes}        &                                                            \\
\textit{Calcidiscus}       & Kamptner, 1950                                             \\
\textit{Calciopappus}      & Gaarder \& Ramsfjell, 1954                                 \\
\textit{Calciosolenia}     & Gran, 1912                                                 \\
\textit{Calyptrosphaera}   & Lohmann, 1902                                              \\
\textit{Coccolithus}       & E.H.L.Schwarz, 1894                                        \\
\textit{Cyclolithella}     & Loeblich \& Tappan, 1963                                   \\
\textit{Emiliania}         & W.W.Hay \& H.P.Mohler, 1967                                \\
\textit{Michaelsarsia}     & Gran, 1912                                                 \\
\textit{Ophiaster}         & Gran, 1912                                                 \\
\textit{Syracosphaera}     & Lohmann, 1902                                              \\
\textit{Umbellosphaera}    & Paasche, 1955                                              \\
\textit{Umbilicosphaera}   & Lohmann, 1902                                              \\
\textbf{Others}            &                                                            \\
\textit{Dictyocha}         & Ehrenberg, 1837                                            \\
\textit{Eutreptia}         & Perty, 1852                                                \\
\textit{Eutreptiella}      & A.M.Cunha, 1913                                            \\
\textit{Mesodinium}        & von Stein                                                 
\end{tabular}
\end{table}
